The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-25 manuscript/manuscript.Rnw
remove comparative evolution section
fix link to crh figure
ditch opossum shift
Completely rewrite conclusion

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}). These genes
include multiple genes which are thought to be involved in
pre-eclampsia. Notably, \textit{CRH}
(FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
involved in modulating trophoblast invasion. Two other genes,
\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

@@

% \textit{Ateles fuciceps} shows enrichment in extracellular
% matrix and structural molecular activity (Supplemental
% \cref{tab:go_terms_enriched_ateles_fusciceps}). \textit{Homo sapiens}
% placentas are enriched in female pregnancy, estrogen biosynthesis,
% reproductive process, hormone aactivity, oxygen transporter activity,
% and haptoglobin binding (Supplemental
% \cref{tab:go_terms_enriched_homo_sapiens}). \textit{Pan paniscus}
% placentas are enriched in clusters related to stress, apoptosis, and
% response to abiotic stimuli. \textit{Mus musculus} placentas are
% enriched in cell junction, vesicle transport, in utero embryonic
% development, and many cytoskeletal processes. \textit{Spalax
%   galili} placentas are enriched in extracellular matrix organization,
% integrin binding, and cell death.
% 
% Cow shows only four
% significantly enriched clusters; however many genes highly expressed
% in the cow are ruminant-specific placenta genes and are not yet
% annotated to specific GO categories. Dog shows enrichment in genes
% associated with RNA, as well as homeostasis, glycoproteins, lyosome,
% and calcium binding. Elephant shows enrichment in genes associated
% with RNA, coagulation and calcium binding. Opossum shows enrichment in
% genes involved in apoptosis, oxidative stress, mitochondrion, hormone
% metabolism, and lipids.

% Not sure if this is actually worthwhile; I've mentioned some of this
% already for some of the lineages

% \subsection{Comparative Analysis of Gene Expression}
% 
% We performed a comparative analysis of gene expression by
% reconstructing the evolution of expression values and detecting
% significant shifts in expression along specific lineages. Likelihood
% values for each model for each of the X genes reconstructed, as well
% as likelihood ratio tests, P- and Q-values (after correcting for
% multiple testing) are shown in Supplemental Table X. We identified
% between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
% 0.01) in expression on specific lineages (Table 10). The top 10 genes
% with significant shifts (if applicable) per lineage are shown in Table
% 10 and all genes for each branch are shown in Supplemental Tables
% X-XX. Genes that showed shifts on the human lineage were
% overwhelmingly identified as placenta-specific by the
% tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
% both TSI>0.75 and placenta as one the top two tissues with the highest
% expression. Unsurpisingly, this was not the case with other lineages,
% since we used expression values from a human-derived microarray to
% generate TSI values for each gene. However, this does underscore the
% dynamic nature of expression patterns in the placenta and indeed all
% tissues across mammals.
% 
% We then performed DAVID functional annotation clustering analyses on
% each list of genes from each lineage. The top three clusters for each
% lineage are shown in Table 10 and a full list of clusters in
% Supplemental Table X. The top enrichment cluster in the human lineage
% is blood vessel development, followed by extracellular matrix and
% metal ion binding (Table 10). Other significant clusters include those
% related to cell communication/signaling, cytokines, defense response,
% tryptophan metabolism, apoptosis, and skeletal system development
% (Supplemental Table X). The bonobo lineage was enriched in only three
% clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
% mouse displayed enrichment in clusters involved in epigenetic
% modifications such as methyltransferase activity and histone
% modification, as well embryonic morphogeneisis, lipid biosynthesis,
% among others (Supplemental Table X). The top enriched cluster in the
% cow lineage was the response to vitamin A, as well as many clusters
% related to the immune system. The dog showed enrichment in skeletal
% system development, as well as lipid metabolic processes and blood
% circulation. The elephant lineage had only one enriched cluster:
% carbohydrate binding. Genes that showed enrichment on the opossum
% lineage (essentially representing the difference between placental
% mammals and marsupials) were clustered in many categories not seen in
% the other lineages, including clusters related to the mitochondria,
% urea cycle, peroxisome, glycolysis, and transport across the cell
% membrane (Table 10 and Supplemental Table X).

@@

% Genes that show a significant shift on the \opossum lineage include
% functional annotation clusters that correspond to active generation of
% energy, transport, and lipid breakdown via the peroxisome (Table 10;
% Supplemental Table X). The choriovitelline placenta of both didelphids
% and the tammar wallaby are known to be involved in glycolysis and
% protein synethsis (Freyer et al., 2003). This implies that the
% choriovitelline placenta in marsupials is an active site of metabolism
% and molecular synthesis, as well as the transport of nutrients from
% the maternal tissues. This is a distinct difference that is not seen
% in significant shifts on other lineages, and may be due to the
% formation of the placenta in the opossum from the yolk sac. In
% addition, the urea cycle is a highly enriched functional annotation
% cluster which directly correlates with the high levels of urea in the
% Monodelphis placenta (Freyer et al., 2007). This is potentially a
% Monodelphis-specific characteristic as discussed above.

@@

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. Although there are speculative cases of
eclampsia in great apes
\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is possible
that a comparative evolutionary approach to gene expression can
identify genes that have divergent expression patterns from other
mammals as well as association with pre-eclampsia.

Functional annotation clusters significantly enriched on the human
lineage (blood vessel development, extracellular matrix, tube
morphogenesis) seem to suggest involvement in remodeling of tissues
and increased blood flow, both of which may correlate with the
invasion of the trophoblast (\cref{tab:go_terms_all_species}). In
addition, clusters such as cytokines and defense response
(\cref{tab:go_terms_all_species}) also indicate an a significant
change in expression of the immune response that may be involved in
immunosupression at the maternal-fetal boundary to further allow
invasion of fetal tissues.

\DLA{Make the discussion of pre-eclampsia clearer and more defined;
  possibly include some of the discuss below. Also, this is not
  necessarily a shift on the human line.}

% I haven't actually done this analysis, so commented out

% Many genes that both show expression shifts on the human lineage and
% have very high expression (FPKM>200) are directly related to the
% highly invasive nature of the human placenta. For example, GPC3,
% glypican 3, is an imprinted gene and cell growth regulator, associated
% with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
% intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
% 1996). GPC3 has the highest expression in humans (Table 2), but is
% also expressed in other mammals to a lesser degree, especially those
% with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
% spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
% (Supplemental Tables X-XX). ADAM12 increases the availability of
% insulin-like growth factors that lead to the promotion of cell growth
% and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
% found evidence of adaptive evolution in ADAM12 on the primate lineage,
% and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
% monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
% lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
% associated with trophoblast invasion as well (Bilban et al., 2010;
% Yuen et al., 2011). In addition, one other gene, KISS1, shows
% exceptional abundance in the human placenta (Table 2); however due to
% its absence in elephant Ensmebl gene annotations, it was not part of
% the 1:1 orthologues examined in the comparative study. Here, it is
% only expressed in the term placentas of humans (Table 2) and bonobos
% (\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
% in the regulation of trophoblast invasion, and shows highest
% expression in the first trimester of pregnancy (Bilban et al., 2004;
% Reynolds et al., 2009).
% 
% In addition, several other genes that show expression shifts on the
% human lineage also have major effects on placental function. Genes
% such as TXK and EBI3 are intimately involved in the immune system in
% the placenta. TXK is involved in natural killer cell differentiation
% (Bezman et al., 2012). EBI3 is implicated in natural killer cell
% function an maternal-fetal immunomodulation in the syncytiotrophoblast
% (Devergne et al., 2001). KYNU and KMO are interacting genes that are
% involved in tryptophan metabolism, which may be involved in
% immunosupression and the prevention of fetal rejection (Takikawa,
% 2005). THSD7A is a thrombospondin, which inhibits endothelial cell
% migration and tube morphogenesis (Wang et al., 2010).
% 
% Many genes that show evidence of a shift on the human lineage show
% evidence of implication in pre-eclampsia, including ADAM12 (Gack et
% al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
% Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
% al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
% MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
% In addition KISS1 is also involved in pre-eclampsia as well (Sado et
% al., 2011; Qiao et al, 2012). Due to the association of many genes
% that show human-specific shifts as well as high expression, we
% conclude that pre-eclampsia is possibly a symptom of recent evolution
% of gene expression on the human lineage. Although many genes such as
% GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
% this expression has intensified on the human lineage. Sequencing of
% more placentas of nonhuman primates, including other individuals of
% the genus Pan will help to further characterize this transition.

@@

\DLA{Totally rewrite the conclusion; it's terrible}

We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
spanning the phylogenetic tree of mammals. This revealed a dynamic
expression pattern of expression of lineage-specific expansions of
gene families and differential gene expression that agreed with the
scenario of an “arms race” between maternal and fetal tissues that
have shaped such disparate gene expression profiles. In addition, we
highlighted the broad spectrum of hormone gene expression in multiple
species that suggested that there is a lack of conservation in hormone
signaling during mammalian pregnancy. Major differences between
expression in marsupial and eutherian mammals at term seemed to be
correlated with the differences between choriovitelline and
chorioamniotic placenta, with the marsupial placenta more enriched in
genes which contribute to the active generation of energy. We also
identified that galectin expression, especially that of
\textit{LGALS1}, is a common feature of all mammalian placentas and
was most likely present in the ancestral therian. Also, we identified
two new expansions of genes that are highly expressed in the \elephant
placenta that may be involved in trophoplast differentiation and immune
system function, mirroring expansions in other groups such as
primates, rodents, and ruminants. We also quantified significant
lineage-specific shifts in gene expression, concentrating on the
identification of genes that showed gene expression changes on the
human lineage. We cautiously propose that many of these genes may be
related to the invasive nature of the trophoblast in \human, although
we note that in many of these genes, there is in some cases
substantial expression in the \bonobo as well. These genes have the
potential to spur further research and may be at the root of
obstetrical disorders such as pre-eclampsia. We propose that more
placentas from diverse lineages, especially those of marsupials and
primates, to further identify genes and expression levels that are
specific to marsupials and eutherians, as well as confirm the identity
of human-specific expression shifts.

>>>>

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}). These genes
include multiple genes which are thought to be involved in
pre-eclampsia. Notably, \textit{CRH}
(FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:crh_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
involved in modulating trophoblast invasion. Two other genes,
\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

@@

@@

@@

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. Although there are speculative cases of
eclampsia in great apes
\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

The identification of multiple genes (\textit{CRH}, \textit{ADAM12},
\textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1}, and \textit{PLAC1};
\cref{tab:human_lineage_specific}) which are associated with
pre-eclampsia which have increased expression on the \human lineage
may explain the larger proportion of pre-eclampsia observed in humans.
However, there are still genes which are associated with pre-eclampsia
which are also more highly expressed in the \mouse lineage such as
\textit{Afp} and \textit{Gjb2}. Furthermore, the onset of
pre-eclampsia is well before term; all of our placentas are term
placentas and provide limited information about the transcriptome
during the onset of pre-eclampsia.

@@

We analyzed the sequenced transcriptome of
\Sexpr{star.logs[,length(unique(file))]} placentas from
\Sexpr{star.logs[,length(unique(species))]} species spanning the
phylogenetic tree of mammals. This wealth of data revealed both
lineage-specific expansions and differential expression of gene
families as well as a set of core, non-housekeeping genes which are
critical for the proper organization and functioning of the placenta.

There is diversity in the placental expression of hormones, including
estrogen, progesterone, prolactin, relaxin, natriuretic peptides, and
the growth hormone family amount the species and lineages analyzed. We
also identified genes differentially expressed on the \human lineage
which are implicated in pre-eclampsia, including \textit{ADAM12},
\textit{CRH}, and \textit{KISS1}.

The core placenta transcriptome was also identified, which contains
genes which are involved in maternal-fetal tolerance, such as the
annexins, and genes responsible for cell-cell and cell-matrix
interactions, such as \textit{LGALS1}. These genes were almost
certainly present in the ancestral therian placenta. We hypothesize
that modifications to their expression or function will compromise the
development and organization of the placenta.

This study has identified many tantalizing molecular pathways which
illuminate the vast diversity of placental morphology and may explain
some placental malfunctions in \human{}. However, the strength of our
hypotheses are limited by our limited data; more placentas from
multiple members of diverse clades and different timepoints are still
needed to come to a more complete understanding of the genomic and
molecular basis for placenta morphology and function. We look forward
to collecting and analyzing such a dataset.
